Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2024-11-03 paper/main.tex
wip addressing reviewers' comments

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage{soul}
\usepackage{enumerate}
\usepackage[table]{xcolor}
\usepackage{authblk}
%%%%%%%%%%%%%%%%%

@@

@@

Conceptualizing strategies and understanding the best way to play the game have
been of interest to the scientific community since the formulation of the game
\cite{Axelrod1980a, Beaufils1997, Press2012, Tzafestas2000, Hilbe2017,
Glynatsi2020, Murase2021, Schmid2022, Chen2023}.
This extends to both tournament settings and population dynamics.
Computer tournaments became a common evaluation technique for newly designed
strategies following Axelrod's computer tournaments in the
1980s~\cite{Axelrod1980a, Axelrod1980b}.
The winner of both of Axelrod's tournaments~\cite{Axelrod1980a, Axelrod1980b}
was the simple strategy Tit For Tat (\TFT{}). \TFT{} cooperates on the first turn
and thereafter copies the previous action of its opponent, retaliating against
defections with a defection and forgiving a defection if followed by
cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions for success in an \IPD{}
tournament:

@@

@@

@@

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[b]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 &
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{\textbf{Result Summary Example of a Tournament.}
A result summary consists of \(N\) rows, with each row containing information
for each strategy that participated in the tournament. This information includes
the strategy's rank (\(R\)), median score, the cooperation rate (\(C_r\)), the number of
match wins, and the probability that the strategy cooperated in the opening
move. Additionally, it provides the probabilities of a strategy being in any of
the four states ($CC, CD, DC, DD$) and the cooperation rate after each state.}\label{table:output_result}
\end{table}

@@

$C_{\text{min}}$ / $C_r$ & A strategy's cooperating rate divided by the minimum & result summary  & float & 0 & 1 \\

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.8\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{\textbf{Correlations between the features of Table~\ref{table:manual_features}
and the normalised rank and the median score.}
The correlation coefficients are calculated using the Spearman's rank
correlation coefficient.}\label{table:correlations}
\end{table}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=\textwidth]{../images/features_result.pdf}
        \caption{\textbf{Distributions of \(C_r\) and \(C_r / C_{\text{mean}}\)
        for the winners of tournaments.} A value of \(C_r / C_{\text{mean}} =
        1\) imply that the cooperating ratio of the winner was the same as the
        mean cooperating ratio of the tournament.}
        \label{fig:discussion_cooperation_measures}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=.75\textwidth]{../images/sse.pdf}
        \caption{\textbf{Distributions of SSE error for the winners of
        tournaments.} As a reminder, the SSE error shows how close a strategy is
        to behaving as a ZDs, and subsequently, in an extortionate way. A SSE
        value of 1 indicates no extortionate behaviour at all whereas a value of
        0 indicates that a strategy is behaving as a ZDs.}
        \label{fig:discussion_sse}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=\textwidth]{../images/rates_result.pdf}
        \caption{\textbf{Distributions of rates $CC$ to $C$, $CD$ to $C$, $DC$
        to $C$, and $DD$ to $C$ for the winners of
        tournaments.}}\label{fig:discussion_rates}
\end{figure}

@@

>>>>

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage{soul}
\usepackage{enumerate}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{listings}
%%%%%%%%%%%%%%%%%

@@

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{codegreen},
    stringstyle=\color{codegreen},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                                             
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

@@

Conceptualizing strategies and understanding the best way to play the game have
been of interest to the scientific community since the formulation of the game
\cite{Axelrod1980a, Nowak1992, Nowak1993, Beaufils1997, Press2012, Tzafestas2000, Hilbe2017,
Glynatsi2020, Murase2021, Schmid2022, Li2022, Chen2023}.
This extends to both tournament settings and population dynamics.
Computer tournaments became a common evaluation technique for newly designed
strategies following Axelrod's computer tournaments in the
1980s~\cite{Axelrod1980a, Axelrod1980b}.
The winner of both of Axelrod's tournaments~\cite{Axelrod1980a, Axelrod1980b}
was the simple strategy Tit For Tat (\TFT{}). \TFT{} cooperates on the first turn
and thereafter copies the previous action of its opponent, retaliating against
defections with a defection and forgiving a defection if followed by
cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions for success in an \IPD{}
tournament:

@@

Beyond these difficulties, we believe that limited comprehensive analyses are
rooted in field conventions. Tournaments or evolutionary dynamics often rely on
a select list of hand-picked strategies chosen by modelers, typically based on
specific properties they wish to examine. This practice may stem from
misconceptions, such as the assumption that because \TFT{} performs
relatively well, it is sufficient to test only against \TFT{} variants.
Another misconception may stem from the Press \& Dyson result~\cite{Press2012},
which implies that lower memory strategies will always dominate pairwise
interactions, leading some to consider only memory-one strategies. However, this
result holds strictly only in pairwise interactions.

It is not only the set of strategies or the tournament parameters such as noise
that may impact results but also the design of the round-robin tournament itself. To
address this~\cite{Rapoport2015} separately examined the effects of changes in format, objective
criteria, and payoff values on tournament outcomes. They demonstrated that
\TFT's performance declined under certain conditions. To our knowledge,
this is the only study that has reanalyzed the tournament structure and
critically evaluated it. However, in this work, we employ an extensive list of
strategies made possible by the \axelrod{} package, an approach that would
have been difficult to achieve previously. Unlike the authors of that study, we
do not consider a new tournament design.

@@

To run a tournament, only a few lines of code are required
(Figure~\ref{fig:axelrod_code}). Specifically, one needs to define the list of
strategies that the players use when participating in the tournament, the number
of repetitions, the number of turns or the probability of each match ending, and
the probability of noise. We demonstrate two examples in
Figure~\ref{fig:axelrod_code}: one with a fixed number of turns and one with a
probabilistic ending. A tournament is an instance of the `Tournament` class. To
execute the tournament, users need to run the `play()` method, which returns an
instance of the `Results` class. This instance includes many details of the
tournament, such as the winner and the average score. Additionally, a more
detailed summary of the tournament is provided, which we use in this analysis
and describe in detail below.

\begin{figure}
\begin{lstlisting}[language=Python]
>>> import axelrod as axl

>>> players = [axl.TitForTat(), axl.GTFT(), axl.Gradual(), axl.EvolvedLookerUp2_2_2()]

>>> standard_tournament = axl.Tournament(players=players, turns=50, repetitions=10, noise=0)

>>> results = standard_tournament.play()

>>> prob_noisy_end_tournament = axl.Tournament(players=players, prob_end=0.5,
...                                            repetitions=10, noise=0.001)

>>> results = prob_noisy_end_tournament.play()
\end{lstlisting}
\caption{{\bf Example usage of the \axelrod package for running tournaments.}
The strategies that players use are saved in a list, which is then passed to the
Tournament class. In this example, strategies we previously discussed, such as \TFT, Generous Tit-for-Tat, Gradual, and a stochastic strategy
evolved through reinforcement learning, are included. We create a standard
tournament where noise is set to 0, as
well as a noisy tournament with probabilistic ending. Once an instance of the
tournament class is defined, executing the tournament is straightforward. The play()
method generates all possible pairs from the list of strategies, including
pairs where each strategy plays against itself, and then iterates through each
match to play it. Each match is repeated according to the specified number of
repetitions, with results aggregated to a tournament summary.}\label{fig:axelrod_code}
\end{figure}

@@

The summary contains statistics regarding each strategy that participated in the
tournament, such as its rank, cooperation rate, time spent in each state, and
the probability of cooperating after each of the four possible outcomes of the
previous round. In our analysis, we will use the measures for each strategy that
are provided in the summary, as well as additional measures we calculated.
Namely, these include the SSE error, the average cooperation in each tournament,
the maximum, and the minimum values. The SSE (introduced in~\cite{Knight2024})
shows how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. We also consider how each strategy's cooperation rate \(C_r\)
compares to those of the tournament as a whole, for example, by comparing
\(C_r\) to \(C_{\text{max}}\).

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[t]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 &
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{\textbf{Result Summary Example of a Tournament.}
A result summary consists of \(N\) rows, with each row containing information
for each strategy that participated in the tournament. This information includes
the strategy's rank (\(R\)), median score, the cooperation rate (\(C_r\)), the number of
match wins, and the probability that the strategy cooperated in the opening
move. Additionally, it provides the probabilities of a strategy being in any of
the four states ($CC, CD, DC, DD$) and the cooperation rate after each state.}\label{table:output_result}
\end{table}

@@

$C_{\text{min}}$ / $C_r$ & The minimum divided by a strategy's cooperating rate & result summary  & float & 0 & 1 \\

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.8\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{\textbf{Correlations between the features of Table~\ref{table:manual_features}
and the normalised rank and the median score.}
For each type of tournament, standard, noisy, probabilistic ending, and noisy
probabilistic ending, we run a correlation analsysif. For each tournament we
check the correlation between each feature used in our analysis and the normalised
random and median score. Note that the correlation coefficients are calculated
using the Spearman's rank correlation coefficient. A negative value
indicates a negative correlation and in the case of the normalised rank
a smaller rank translats to better position in the tournament.}\label{table:correlations}
\end{table}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=\textwidth]{../images/features_result.pdf}
        \caption{{\bf Distributions of \(C_r\) and \(C_r / C_{\text{mean}}\)
        for the winners of tournaments.} In this distribution, we consider the
        winners of the tournaments, specifically the strategies that ranked
        first in each tournament. For each type of tournament, we plot the
        cooperation rate of the winner in the tournament they won, as well as
        the ratio of the winner's cooperation rate to that of the entire
        tournament. A value of \(C_r / C_{\text{mean}} = 1\) implies that the
        cooperation ratio of the winner was the same as the mean cooperation
        ratio of the tournament.}
        \label{fig:discussion_cooperation_measures}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=.75\textwidth]{../images/sse.pdf}
        \caption{\textbf{Distributions of SSE error for the winners of tournaments.}
        Here, we again consider the winners of the tournaments, separated by type of
        tournament, and plot their SSE error. As a reminder, the SSE error indicates how
        closely a strategy behaves like a Zero-Determinant (ZD) strategy, and
        subsequently, in an extortionate way. An SSE value of 1 indicates no
        extortionate behavior at all, whereas a value of 0 indicates that a strategy is
        behaving as a ZD.}
        \label{fig:discussion_sse}
\end{figure}

@@

\begin{figure}[!htbp]
    \centering
        \centering
        \includegraphics[width=\textwidth]{../images/rates_result.pdf}
        \caption{\textbf{Distributions of rates \(CC\) to \(C\), \(CD\) to \(C\), \(DC\)
        to \(C\), and \(DD\) to \(C\) for the winners of tournaments.} The result
        summary from the tournaments records how often each strategy cooperated after
        each possible outcome of the previous round. Specifically, we analyze the
        probability with which a strategy chose \(C\) following the outcomes. Here, we
        plot the distributions of these probabilities for the winners of the
        tournaments. We separate them by type of tournament, from top to bottom:
        standard, noisy, probabilistic ending, and noisy probabilistic ending. From left
        to right, we plot the distributions of cooperation after the outcomes \(CC\),
        \(CD\), \(DC\), and \(DD\).}\label{fig:discussion_rates}
\end{figure}

@@

To this end, many strategies, such as Win-Stay-Lose-Shift and Generous
Tit For Tat, emerged due to their strong performance in evolutionary dynamics.
Axelrod's original work relied on computer tournaments, so we chose to remain
consistent with this approach, as a comprehensive study like ours had not yet
been undertaken. However, evolutionary settings would be an exciting direction
for future study.
